The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-11 rational.tex
sat

The construction presented here is a novel SAT-solver that relies on
{{trace*}} for its solution. It varies from previous classical SAT solvers
\amr{must cite ???} in that there is no explicit search operation on the
boolean space. It also varies from previous quantum SAT solvers \amr{must
  cite ???} in that it does uses a {{trace}} operation to compute a fixpoint
which is the solution following which an isomorphic clone operation allows us
to examine the result.  Though we focus on SAT here, this construction can be
generalized to any constraint satisfaction problem whose search space can be
represented by a recursive type and for which an isomorphic clone operation
can be constructed. Before we start, let us recall two constructions:

\begin{enumerate}
\item Recall from previous work that for the types {{bool}} and, by
  extension, any tuple of {{bool}}, an isomorphic clone operation can
  be constructed (See Lemma 7.2 \cite{infeffects}).

\item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
  {{if_{not} }}, which when generalized to two bits was
  {{toffoli}}. This construction maybe generalised to {{n}} bits as
  {{if^n_{not} }}.
\end{enumerate}

@@

\emph{Controlling the heap.}  The final detail we have to be concerned
with is the handling of the heap and garbage of {{f}}. Any boolean
expression {{bool^n -> bool}} can be compiled into the isomorphism
{{f:bool^h*bool^n<->bool^g*bool}} where the extra bits {{bool^h}} and
{{bool^g}} are the heap and garbage. Constructing such an isomorphism
has been detailed before \cite{Toffoli:1980,infeffects}.  However the
computation of {{f}} is meaningful {{iff}} the heap {{bool^h}} is
initialized appropriately. We can ensure that the heap has the
appropriate initial value by checking the heap and negating a second
control wire, if the values do not match, i.e., the dotted part in
the diagram below and is {{if^h_{not} }} followed by a {{not}} on the
`heap control wire'.

@@

\emph{Cloning the Solution.}.  Let us call the above construction
which maps inputs, heap and control wires to input, heap and control
wires as {{sat_f}}. The SAT-solver is is completed by tracing the
{{sat_f}} and cloning the `inputs'.

>>>>

The construction presented here is a novel SAT-solver that relies on
{{trace*}} for its solution. It varies from previous classical SAT solvers
\amr{must cite ???} in that there is no explicit search operation on the
boolean space. It also varies from previous quantum SAT solvers \amr{must
  cite ???} in that it does uses a {{trace}} operation to compute a fixpoint
which is the solution following which an isomorphic clone operation allows us
to examine the result.  Though we focus on SAT here, this construction can be
generalized to any constraint satisfaction problem whose search space can be
represented by a recursive type and for which an isomorphic clone operation
can be constructed. Before we start, let us recall a constructions from 
prior work \cite{infeffects}.

\emph{Conditionals.}
Given any combinator {{c : b <-> b}} we can construct a combinator called
{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
the combinator {{c}} is used to transform the value of type~{{b}}. If the
boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
write down the combinator for {{if_c}} in terms of {{c}} as 
{{ dist (;) ((id (*) c) (+) id) (;) factor }}.

% Let us look at the combinator pictorially as if it were a circuit and values
% are like particles that flow through this circuit.

The diagram below shows the input value of type {{(1+1)*b}} processed
by the distribute operator {{dist}}, which converts it into a value of
type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
case when the boolean is {{true}} (i.e. the value was {{left ()}}),
the combinator~{{c}} is applied to the value of type~{{b}}. The right
branch which corresponds to the boolean being {{false}} passes along
the value of type {{b}} unchanged.

\begin{center}
\scalebox{1.0}{
%%subcode-line{pdfimage}[diagrams/if_c.pdf]
\includegraphics{diagrams/thesis/cnot.pdf}
}
\end{center}

% We will be seeing many more such wiring diagrams in this paper and it is
% useful to note some conventions about them. Wires indicate a
% value that can exist in the program. Each wire, whenever possible, is
% annotated with its type and sometimes additional information to help clarify
% its role. When multiple wires run in parallel, it means that those values
% exist in the system at the same time, indicating pair types. When there is a
% disjunction, we put a {{+}} between the wires. 
% Combinators for distribution {{dist}} and factoring {{factor}}
% are represented as triangles with their operator symbols in them. Other
% triangles may be used and, in each case, types or labels will be used to
% clarify their roles. Finally, we don't draw boxes for combinators such as
% {{id}}, commutativity, and associativity, but instead just shuffle the wires
% as appropriate.

\emph{Boolean Operation based on {{if_c}}.}
The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
negates its second argument if the first argument is {{true}}. This
gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
equivalent construction that is useful is {{else_{not} }} where we
negate the second argument only is the first is {{false}}.

Similarly, we can iterate the construction of {{if_c}} to check
several bits. The gate {{if_{cnot} }}, which we may also write as
{{if^2_{not} }}, checks two booleans and negates the result wire only
if they are both {{true}}. The gate {{if^2_{not} }} is well known as
the universal reversible primitive gate called the `Toffoli gate'
\cite{Toffoli:1980}.  We can generalize this construction to
     {{if^n_{not} }} which checks {{n}} bits and negates the result
     wire only if they are all {{true}}.

\emph{Cloning Bits.} The gate {{else_{not} }} has the interesting
property that as long as the second argument is fixed to be {{true}},
output of the second wire is always the same as teh value of the first
wire. In other words {{else_{not} }} can be thought of as cloning the
value of the first wire. A circuit of {{n}} parallel {{else_{not} }}
gates can hence clone {{n}} bits. They also consume {{n}} {{true}}
inputs in the process. Let us call this construction 
{{clone^n_{bool} }}.

%% \begin{enumerate}
%% \item Recall from previous work that for the types {{bool}} and, by
%%   extension, any tuple of {{bool}}, an isomorphic clone operation can
%%   be constructed (See Lemma 7.2 \cite{infeffects}).

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
%%   {{if_{not} }}, which when generalized to two bits was
%%   {{toffoli}}. This construction maybe generalised to {{n}} bits as
%%   {{if^n_{not} }}.
%% \end{enumerate}

@@

\emph{Controlling the heap.}  The final detail we have to be concerned
with is the handling of the heap and garbage of {{f}}. Any boolean
expression {{bool^n -> bool}} can be compiled into the isomorphism
{{f:bool^h*bool^n<->bool^g*bool}} where the extra bits {{bool^h}} and
{{bool^g}} are the heap and garbage. Constructing such an isomorphism
has been detailed before \cite{Toffoli:1980,infeffects}.  However the
computation of {{f}} is meaningful {{iff}} the heap {{bool^h}} is
initialized appropriately. We can ensure that the heap has the
appropriate initial value by checking the heap and negating a second
control wire, if the values do not match, i.e., the dotted part in the
diagram below and is {{if^h_{not} }} followed by a {{not}} operator on
the control wire.

@@

\emph{Cloning the Solution.}.  Let us call the above construction
which maps inputs, heap and control wires to input, heap and control
wires as {{sat_f}}. The SAT-solver is is completed by tracing the
{{sat_f}} and cloning the `inputs' using {{clone^n_{bool} }}.
